---
title: "scGAS: Gene Activation Potential from scATAC-seq"
author: "Weiyan Chen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{scGAS Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

**scGAS** infers Gene Activation Potential (GAS) from scATAC-seq data. Starting
from raw fragment files, it produces per-gene, per-cell activation scores and a
Chromatin Potential Field that predicts differentiation trajectories.

Every function from Step 2 onwards takes a `SeuratObject` as its first
argument and returns the same object enriched with new results — either as cell
metadata, a new assay, or an entry in `@misc`.  Step 5 is the only exception:
it returns a standalone `cpf` object.

| Step | Function | What is added |
|------|----------|---------------|
| 1 | `scgas_preprocess()` | creates `SeuratObject` (ATAC assay, LSI, UMAP, KNN graph) |
| 2 | `scgas_metacell()` | `obj$mc_membership` + `obj@misc$mc_membership` |
| 3 | `scgas_train_models()` | `obj@misc$scgas_models` |
| 4 | `scgas_compute()` | `scGAS` assay, `scgaspca` and `scgasumap` reductions |
| 5 | `scgas_chromatin_potential()` | returns a `cpf` object |

---

## Installation

```{r install}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "Signac", "GenomicRanges", "rtracklayer",
  "EnsDb.Hsapiens.v75", "SCAVENGE"
))

install.packages(c("Seurat", "glmnet", "glmnetUtils", "igraph", "ggplot2"))

remotes::install_github("ChenWeiyan/scGAS", subdir = "scGAS-R")
```

---

## Required reference files

Reference files (~570 MB for hg19) are hosted on Zenodo.  Download them once
after installing the package:

```{r download}
## Downloads all hg19 reference files into ./data/
scgas_download_reference(genome = "hg19", data_dir = "data")

## hg38 (when available):
# scgas_download_reference(genome = "hg38", data_dir = "data")
```

| File | Size | Description |
|------|------|-------------|
| `hg19_500bp_CRE.bed` | 68 MB | Reference cCRE BED (1.8 M × 500 bp bins) |
| `ENCODE_167Tissue_SigAss_pVal.05.gz` | 30 MB | cCRE–Gene association list (p < 0.05) |
| `ENCODE_167Tissue_RNA_LogNorm.gz` | 13 MB | ENCODE bulk RNA-seq, log-normalised |
| `ENCODE_167Tissue_CRE_LogNorm.gz` | 448 MB | ENCODE bulk DNase-seq, log-normalised |

---

## Example dataset

This vignette uses the 10x Genomics 10k Human Brain Multiome (GEMx) dataset.

```{r paths}
fragment_file <- "datasets/10k_Human_Brain_MO_gemx_atac_fragments.tsv.gz"
data_dir      <- "data"   # contains hg19_500bp_CRE.bed, ENCODE *.gz files
```

---

## Step 0 — Gene annotations

`scgas_preprocess()` detects the genome assembly from the fragment file header
(written by CellRanger / CellRanger-ARC) and sets both `seqlevelsStyle` and
`genome` on the annotation automatically.  You only need `GetGRangesFromEnsDb()`.

```{r annotations}
library(EnsDb.Hsapiens.v75)
library(Signac)

## No seqlevelsStyle() or genome() call needed — handled by scgas_preprocess()
annotation <- GetGRangesFromEnsDb(EnsDb.Hsapiens.v75)
```

---

## Step 1 — Preprocess

`scgas_preprocess()` handles everything from raw fragment files to a
dimensionality-reduced `SeuratObject`.

```{r preprocess}
library(scGAS)

## Genome is auto-detected from the fragment header.
## CRE BED and ENCODE reference files are found automatically in data_dir.
obj <- scgas_preprocess(
  fragment_path     = fragment_file,
  data_dir          = data_dir,
  annotation        = annotation,
  lsi_dims          = 2:30,
  umap_n_components = 3L,
  n_cores           = 8
)

## genome_version and data_dir are stored in obj@misc for downstream steps
obj@misc$genome_version   # e.g. "hg38" or "hg19"
obj@misc$data_dir

## Override genome detection if needed (skips auto-detection entirely)
# obj <- scgas_preprocess(..., genome_version = "hg19")
```

Inspect QC before filtering:

```{r qc}
library(Seurat)
VlnPlot(obj,
        features = c("nFeature_ATAC", "TSS.enrichment", "nucleosome_signal"),
        ncol = 3, pt.size = 0)
```

Filter low-quality cells if needed:

```{r filter}
obj <- subset(
  obj,
  subset = nFeature_ATAC > 200 &
           TSS.enrichment > 1   &
           nucleosome_signal < 4
)
```

---

## Step 2 — Construct Metacells

`scgas_metacell()` reads the `ATAC_snn` graph from the `SeuratObject`, runs
walktrap community detection, and writes `mc_membership` back into
`obj$mc_membership`.

```{r metacell}
obj <- scgas_metacell(
  obj,
  graph_name = "ATAC_snn",
  gamma      = 20,
  min_size   = 10,
  max_size   = 100,
  n_cores    = 8
)

## Inspect Metacell assignments
head(obj$mc_membership)
table(obj$mc_membership)
```

---

## Step 3 — Train Lasso models

`scgas_train_models()` reads `obj$mc_membership` automatically, and loads the
ENCODE reference files from `obj@misc$data_dir` / `obj@misc$genome_version`
automatically.  No file paths needed.

```{r train}
## Simplest call — everything resolved from obj@misc
obj <- scgas_train_models(obj, n_cores = 8)

message(length(obj@misc$scgas_models), " genes modelled")

## Override a reference explicitly if needed
# sig_ass <- readRDS("data/my_custom_sig_ass.rds")
# obj <- scgas_train_models(obj, sig_associations = sig_ass, n_cores = 8)
```

---

## Step 4 — Compute single-cell GAS

`scgas_compute()` reads `obj$mc_membership` and `obj@misc$scgas_models`
automatically.  It adds a `scGAS` assay and `scgaspca` / `scgasumap`
reductions to the returned object.

```{r compute}
obj <- scgas_compute(
  seurat_obj        = obj,
  lsi_dims          = 2:30,
  knn_k             = 30,
  seed_fraction     = 0.10,
  gamma             = 0.05,
  run_dim_reduction = TRUE,
  n_pcs             = 30,
  n_neighbors       = 30,
  n_cores           = 8
)

## Inspect the scGAS assay
dim(obj@assays$scGAS)    # genes x cells
range(obj@assays$scGAS@data)   # 0 – 1
```

Visualise marker gene activation scores:

```{r feature_plot}
FeaturePlot(
  obj,
  features  = c("SLC17A7", "GAD1", "MBP", "AQP4"),
  reduction = "scgasumap",
  ncol      = 2
)
```

---

## Step 5 — Chromatin Potential Field

`scgas_chromatin_potential()` returns a standalone `cpf` object that contains
the arrow data, the ggplot, the HVG matrices, the embedding, and the
parameters used — all independent of the `SeuratObject`.

```{r cpf}
cpf_result <- scgas_chromatin_potential(
  seurat_obj   = obj,
  n_hvg        = 2000,
  knn_k        = 10,
  n_cores      = 8,
  arrow_scale  = 1.5,
  arrow_colour = "steelblue",
  point_colour = "mc_membership",
  title        = "Human Brain — Chromatin Potential Field"
)

## cpf object summary
print(cpf_result)
summary(cpf_result)

## Show the plot
plot(cpf_result)

## Or access directly
print(cpf_result$plot)
```

Subset arrows by magnitude to highlight the most strongly directed cells:

```{r cpf_strong}
library(ggplot2)

strong_df <- cpf_result$cpf_df[
  cpf_result$cpf_df$magnitude > quantile(cpf_result$cpf_df$magnitude, 0.75), ]

ggplot() +
  geom_point(
    data    = cpf_result$cpf_df,
    mapping = aes(x = x, y = y),
    colour  = "grey85", size = 0.3
  ) +
  geom_segment(
    data    = strong_df,
    mapping = aes(x = x, y = y, xend = x + dx, yend = y + dy),
    arrow   = grid::arrow(length = grid::unit(0.10, "cm"), type = "closed"),
    colour  = "firebrick", alpha = 0.8, linewidth = 0.4
  ) +
  theme_classic() +
  labs(title = "Top-quartile CPF arrows", x = "UMAP 1", y = "UMAP 2")
```

Access the underlying data for custom analyses:

```{r cpf_data}
## HVGs used
head(cpf_result$hvg)

## scGAS values for HVGs (genes x cells)
dim(cpf_result$scgas_hvg)

## RNA values for HVGs (genes x cells)
dim(cpf_result$rna_hvg)

## 2-D embedding coordinates
head(cpf_result$embedding)

## Cell metadata
head(cpf_result$cell_metadata)
```

---

## Session information

```{r session, eval=TRUE}
sessionInfo()
```
